---
title: "Visualise longitudinal data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create longitudinal plots to visualise data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ["references.bib"]
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
# Load packages
library(lcsm)
library(ggplot2)
library(tidyr)
library(dplyr)
library(stringr)
```

It's important to understand the data.
For longitudinal analyses it may be helpful visualise the individual trajectories.
A more detailed overview of the rational behind this be found in @Ghisletta2012 or Chapter 2 in @Grimm2017.
Most basic (and advanced) plots can be done relatively easily using already existing R packages.
THe plotting functions of the `lcsm` package actually build on function from other packages.
I'll show some examples illustrating how to visualise longitudinal data using the R package `ggplot2` [@R-ggplot2].
The function `plot_trajectories()` of the `lcsm` R package builds on `ggplot2` to make it a little easier to visualize individual trajectories.

I'll show how to visualise 6 repeated measurements from the data set `data_bi_lcsm` from this package.

# Prepare data

When working with repeated measures I like to create a vector of the variables so I don't have to type the names again and again, but this really needed and everything would also work without doing this.

```{r}
# Create a vector of variable names
x_var_list <- c("x1", "x2", "x3", "x4", "x5", "x6")

# Or simply use the paste function and R might work too
paste0("x", 1:6)
```

The repeated measurements need to be in "long" format when using `ggplot2` for plotting.
Note that this is different to the data structure needed for `lavaan`, so the same data can't just be used for plotting without restructuring - `lavaan` expects the data to be in "wide" format.

I'll show how to transform the data anyway because the plotting function `plot_trajectories()` is limited and other plots may be more appropriate.
Fortunately, this is a relatively straightforward to do, my favourite function to do this is `pivot_longer()` from the `tidyr` package.
Next, to get the correct order of repeated measures the time variable needs to be ordered in the correct order.
R know how to order numbers but only if the variable is actually numeric.
If you're using variable names that also include letters R might get confused and get the order wrong.
Imagine you have four repeated measurements, week 1 to 3 and week 10 and the following variable names: `w1`, `w2`, `w3`, `w10`.
R will get this order wrong and thinks that `w1` is followed by `w10` and then `w2` and `w3`.
To avoid this there are a couple options, I usually do one of the two:

1. Create a factor with the original variable names using the in build `factor()` function from base R
1. Extracting the number from the string and then creating a factor (see example below)

```{r, fig.width=7}
# Create long data set
data_long <- data_bi_lcsm %>% 
  select("id", all_of(x_var_list)) %>% 
  # Pivot data long
  pivot_longer(cols = all_of(x_var_list), names_to = "time", values_to = "value") %>% 
  mutate(
    # Extract number from time variable
    time = str_extract(time, "\\d+"),
    # At the moment the numbers in the time are 'characters'
    # So here it gets transformed to a numeric value
    time = factor(as.numeric(time))
    )
```

These data manipulations are necessary for plotting longitudinal data in R and the `plot_trajectories()` function from this package is doing this transformation automatically in the background.

# Violin plots

```{r, fig.width=7}
# Create violin plot  with outliers in colour blue
# Also add boxplot
ggplot(data_long, aes(time, value)) +
  geom_violin() +
  geom_boxplot(width = 0.1, outlier.colour = "blue") +
  theme_classic()

```

Similar visualisations that communicate the data in a more open way have been developed for example by @Allen2019 and @vanLangen2020.

# Longitudinal plots

While the violin plot focuses on more on the overall distribution, the following plots highlight the individual trajectories for each case in the data.
Longitudinal data can be visualised using the `plot_trajectories()` function from the `lcsm` package.
Here only 1.8% of the data is visualised using the argument `random_sample_frac = 0.018`.
Only consecutive measures are connected by lines as specified in `connect_missing = FALSE`.

## Overlaid individual trajectories

```{r, fig.width=7}
# Create longitudinal plot for construct x
# Select ransom 1.8% of the sample
plot_trajectories(data = data_bi_lcsm,
                  id_var = "id", 
                  var_list = x_var_list,
                  xlab = "Time", ylab = "Value",
                  connect_missing = FALSE, 
                  random_sample_frac = 0.018, 
                  title_n = TRUE)


```

## Separate individual trajectories

```{r, fig.height=5, fig.width=7}
# Create plot for construct x
# Add facet_wrap() function from ggplot2
plot_trajectories(data = data_bi_lcsm,
                  id_var = "id", 
                  var_list = x_var_list,
                  xlab = "Time", ylab = "Value",
                  connect_missing = F, 
                  random_sample_frac = 0.018, 
                  title_n = T) +
  facet_wrap(~id)

```

# References