---
title: "Create path diagrams to visualise model specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create path diagrams to visualise model specifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lcsm)
```

### Plot simplified path diagrams of LCS models

This function is work in progress and can only plot univariate and bivariate LCS models that were specified with `fit_uni_lcsm()` or `fit_bi_lcsm()`.
Modified LCS models will probably return errors as the layout matrix that gets created by this plot function only supports the specifications that can be modelled with this package.
The input arguments for plotting a simplified path dioagram are: 

- the estimated lavaan object `lavaan_object`,
- the `lavaan_syntax` and ,
- `lcsm` indicating whether the LCS model is "univariate" or "bivariate"

Optional arguments can be used to change the look of the plot, for example:

- `lcsm_colours` can be used to highlight the different parts of the latent change score model
  - white: observed scores
  - green: latent true scores
  - blue: latent change scores
  - yellow: latent change scores

Further arguments can be passed on to `semPlot::semPaths()`, for example:

  - `what`, **"path"** to show unweighted gray edges, **"par"** to show parameter estimates as weighted (green/red) edges   
  - `whatLabels`, **"label"** to show edege names as label or **"est"** for parameter estimates, **"hide"** to hide edge labels

#### Univariate LCS model

```{r, fig.width = 6, fig.height = 4}
# Fit bivariate lcsm and save the results 
uni_lavaan_results <- fit_uni_lcsm(data = data_uni_lcsm, 
                                   var = c("x1", "x2", "x3", "x4", "x5"),
                                   model = list(alpha_constant = TRUE, 
                                                beta = FALSE, 
                                                phi = TRUE)
                                  )

# Save the lavaan syntax that is used to create the layout matrix for semPlot
uni_lavaan_syntax <- fit_uni_lcsm(data = data_uni_lcsm, 
                                  var = c("x1", "x2", "x3", "x4", "x5"),
                                  model = list(alpha_constant = TRUE, 
                                               beta = FALSE, 
                                               phi = TRUE),
                                  return_lavaan_syntax = TRUE)

# Plot the results
plot_lcsm(lavaan_object = uni_lavaan_results,
          lavaan_syntax = uni_lavaan_syntax,
          edge.label.cex = .9,  
          lcsm_colours = TRUE,
          lcsm = "univariate")

```

#### Bivariate LCS model

```{r, fig.width = 6, fig.height = 4}
# Fit bivariate lcsm and save the results 
bi_lavaan_results <- fit_bi_lcsm(data = data_bi_lcsm, 
                                 var_x = c("x1", "x2", "x3", "x4", "x5"),
                                 var_y = c("y1", "y2", "y3", "y4", "y5"),
                                 model_x = list(alpha_constant = TRUE, 
                                                beta = TRUE, 
                                                phi = FALSE),
                                 model_y = list(alpha_constant = TRUE, 
                                                beta = TRUE, 
                                                phi = TRUE),
                                 coupling = list(delta_lag_xy = TRUE, 
                                                 xi_lag_yx = TRUE))

# Save the lavaan syntax that is used to create the layout matrix for semPlot
bi_lavaan_syntax <- fit_bi_lcsm(data = data_bi_lcsm, 
                                var_x = c("x1", "x2", "x3", "x4", "x5"),
                                var_y = c("y1", "y2", "y3", "y4", "y5"),
                                model_x = list(alpha_constant = TRUE, 
                                               beta = TRUE, 
                                               phi = FALSE),
                                model_y = list(alpha_constant = TRUE, 
                                               beta = TRUE, 
                                               phi = TRUE),
                                coupling = list(delta_lag_xy = TRUE, 
                                                xi_lag_yx = TRUE),
                                return_lavaan_syntax = TRUE)

# Plot the results
plot_lcsm(lavaan_object = bi_lavaan_results, 
          lavaan_syntax = bi_lavaan_syntax,
          lcsm_colours = TRUE,
          whatLabels = "hide",
          lcsm = "bivariate")

```
